A Model for Extracting Time Delays from Noisy Time Series Data

### 2013-08-27 doc/Paper1/ms.tex
changed kappa notation

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay of image 2	&	Cauchy$\left(0, 0.1\tau_{\rm range}\right)T\left(\tau_{\rm min},\tau_{\rm max}\right)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform$(m_{\rm min}, m_{\rm max})$\\
$\beta$		&	Size of QSO variations		& LogUniform$(\beta_{\rm min}, \beta_{\rm max})$\\
$L_q$		&	Length scale of QSO variations	& LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\{S_1, S_2\}$	&	Size of image-specific variations	& LogUniform$\left(A_{\rm min}, A_{\rm max}\right)$\\
$\{\ell_1, \ell_2\}$ &	Length scale of image-specific variations & LogUniform$(\ell_{\rm min}, \ell_{\rm max})$\\
$\alpha$	&	Smoothness of image-specific variations	& Uniform$(1, 2)$\\
$f_{\rm bad}$	&	Fraction of ``bad'' data points	& Uniform$(0,1)$\\
$\kappa_1$	&	Error-bar boost for ``good'' data & $\frac{1}{2}\delta(\kappa_1 - 1) + \frac{1}{2}$LogUniform$(1,100)$\\
$\kappa_2$	&	Error-bar boost for ``bad'' data & LogUniform$(\kappa_1, 100\kappa_1)$\\
$\{b_i\}_{i=1}^N$ &	Badness hidden variables for each data point & Uniform$(0,1)$\\
\hline
{\bf Derived Parameter} & {\bf Meaning}	&	{\bf Definition}\\
\hline
$S_{q}$	&	Standard deviation of QSO variability	& $\beta\sqrt{L}$\\
$\{a_i\}_{i=1}^N$	&	Errorbar multiplier for each data point & $\left\{\begin{array}{lr}\kappa_1, & b_i > f_{\rm bad}\\ \kappa_2, & b_i \leq f_{\rm bad}\end{array}\right.$\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\caption{A full list of parameters, prior information, and data in the
FLOTSAM model. If there are more than two images, extra parameters are
added accordingly.\label{tab:prior}}
\end{table*}

@@

A common situation in astronomy is that final data products, such as light
curves, are returned, with error bars, from a pipeline. However, such error
bars are not always trustworthy, as they may have been underestimated due to
modelling assumptions and approximations in the pipeline. In the context of
light curve modelling, underestimated error bars can be detected if the
scatter in the light curve is better modelled by independent Gaussian noise
than it is by a time-correlated microlensing term. Thus, to improve the
robustness of our method, we include a ``noise boost'' parameter $\kappa$
describing the factor by which the reported error bars should be multiplied.
This is another mechanism for accounting for outliers.

@@

\section{GP Covariance Function}
Recall that the noise-free light curves of the two images are given by:
\begin{eqnarray}
y_1(t) &=& m_1 + q(t - \tau_1) + \mu_1(t) \\
y_2(t) &=& m_2 + q(t - \tau_2) + \mu_2(t)
\end{eqnarray}
where $\tau_1 = 0$.
Using this, and the model assumptions discussed in the paper, we can obtain the
covariance of image $i$ at time $t_1$ with image $j$ at time $t_2$, where
$i \neq j$. Here, $t_1$ and $t_2$ stand for any two times.
\begin{eqnarray}
\textnormal{Cov}\left[y_i(t_1), y_j(t_2)\right]
&=& \mathds{E}\left[\left(y_i(t_1) - m_i\right)\left(y_j(t_2) - m_j\right)\right]\\
&=& \mathds{E}\left[\left(q(t_1 - \tau_i) + \mu_i(t_1)\right)
\left(q(t_2 - \tau_j) + \mu_j(t_2)\right)\right]\\
&=& \mathds{E}\left[q(t_1 - \tau_i)q(t_2 - \tau_j) + q(t_1 - \tau_i)\mu_j(t_2)
+ \mu_i(t_1)q(t_2 - \tau_j) + \mu_i(t_1)\mu_j(t_2)
\right]\\
&=& S_q\exp
\left[-\frac{|(t_2 -\tau_j)- (t_1-\tau_i)|}{L_q}\right].
\end{eqnarray}
We can also obtain the covariance of image $i$ measured at two times $t_1$ and
$t_2$:
\begin{eqnarray}
\textnormal{Cov}\left[y_i(t_1), y_i(t_2)\right]
&=& \mathds{E}\left[\left(y_i(t_1) - m_i\right)\left(y_i(t_2) - m_i\right)\right]\\
&=& \mathds{E}\left[\left(q(t_1 - \tau_i) + \mu_i(t_1)\right)
\left(q(t_2 - \tau_i) + \mu_i(t_2)\right)\right]\\
&=& \mathds{E}\left[q(t_1 - \tau_i)q(t_2 - \tau_i) + q(t_1 - \tau_i)\mu_i(t_2)
+ \mu_i(t_1)q(t_2 - \tau_i) + \mu_i(t_1)\mu_i(t_2)
\right]\\
&=& S_q\exp
\left[-\frac{|(t_2 -\tau_j)- (t_1-\tau_i)|}{L_q}\right] +
S_i\exp\left[
-\left(\frac{|t_2 - t_1|
}{\ell_i}\right)^\alpha
\right]
.
\end{eqnarray}
The covariance between a data point $Y_i$ and another data point $Y_j$ is
given by the covariance of the underlying noise-free light curves at times $t_i$
and $t_j$, plus a noise term:
\begin{eqnarray}
\textnormal{Cov}\left(
Y_i, Y_j
\right) = \textnormal{Cov}\left(y_{k_i}(t_i), y_{k_j}(t_j)\right)
+ \delta_{ij}\left(a_i\sigma_i\right)^2
\end{eqnarray}
where $\sigma_i$ is the reported error bar given with the data. The coefficient
$a_i$
in front of $\sigma_i$ is the robustness feature that depends on $\kappa_1$,
$\kappa_2$, and the badness variable for data point $i$.

>>>>

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay of image 2	&	Cauchy$\left(0, 0.1\tau_{\rm range}\right)T\left(\tau_{\rm min},\tau_{\rm max}\right)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform$(m_{\rm min}, m_{\rm max})$\\
$\beta$		&	Size of QSO variations		& LogUniform$(\beta_{\rm min}, \beta_{\rm max})$\\
$L_q$		&	Length scale of QSO variations	& LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\{S_1, S_2\}$	&	Size of image-specific variations	& LogUniform$\left(A_{\rm min}, A_{\rm max}\right)$\\
$\{\ell_1, \ell_2\}$ &	Length scale of image-specific variations & LogUniform$(\ell_{\rm min}, \ell_{\rm max})$\\
$\alpha$	&	Smoothness of image-specific variations	& Uniform$(1, 2)$\\
$f_{\rm bad}$	&	Fraction of ``bad'' data points	& Uniform$(0,1)$\\
$\kappa_1$	&	Error-bar boost for ``good'' data & $\frac{1}{2}\delta(\kappa_1 - 1) + \frac{1}{2}$LogUniform$(1,100)$\\
$\kappa_2$	&	Extra error-bar boost for ``bad'' data & LogUniform$(1, 100)$\\
$\{b_i\}_{i=1}^N$ &	Badness hidden variables for each data point & Uniform$(0,1)$\\
\hline
{\bf Derived Parameter} & {\bf Meaning}	&	{\bf Definition}\\
\hline
$S_{q}$	&	Standard deviation of QSO variability	& $\beta\sqrt{L}$\\
$\{a_i\}_{i=1}^N$	&	Errorbar multiplier for each data point & $\left\{\begin{array}{lr}\kappa_1, & b_i > f_{\rm bad}\\ \kappa_1\kappa_2, & b_i \leq f_{\rm bad}\end{array}\right.$\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\caption{A full list of parameters, prior information, and data in the
FLOTSAM model. If there are more than two images, extra parameters are
added accordingly.\label{tab:prior}}
\end{table*}

@@

A common situation in astronomy is that final data products, such as light
curves, are returned, with error bars, from a pipeline. However, such error
bars are not always trustworthy, as they may have been underestimated due to
modelling assumptions and approximations in the pipeline. In the context of
light curve modelling, underestimated error bars can be detected if the
scatter in the light curve is better modelled by independent Gaussian noise
than it is by a time-correlated microlensing term. Thus, to improve the
robustness of our method, we include ``noise boost'' parameters $\kappa_1$,
$\kappa_2$, and $f_{\rm bad}$,
describing the factor by which the reported error bars should be multiplied.
This is another mechanism for accounting for outliers.

@@

\section{GP Covariance Function}
Recall that the noise-free light curves of the two images are given by:
\begin{eqnarray}
y_1(t) &=& m_1 + q(t - \tau_1) + \mu_1(t) \\
y_2(t) &=& m_2 + q(t - \tau_2) + \mu_2(t)
\end{eqnarray}
where $\tau_1 = 0$.
Using this, and the model assumptions discussed in the paper, we can obtain the
covariance of image $i$ at time $t_1$ with image $j$ at time $t_2$, where
$i \neq j$. Here, $t_1$ and $t_2$ stand for any two times.
\begin{eqnarray}
\textnormal{Cov}\left[y_i(t_1), y_j(t_2)\right]
&=& \mathds{E}\left[\left(y_i(t_1) - m_i\right)\left(y_j(t_2) - m_j\right)\right]\\
&=& \mathds{E}\left[\left(q(t_1 - \tau_i) + \mu_i(t_1)\right)
\left(q(t_2 - \tau_j) + \mu_j(t_2)\right)\right]\\
&=& \mathds{E}\left[q(t_1 - \tau_i)q(t_2 - \tau_j) + q(t_1 - \tau_i)\mu_j(t_2)
+ \mu_i(t_1)q(t_2 - \tau_j) + \mu_i(t_1)\mu_j(t_2)
\right]\\
&=& S_q\exp
\left[-\frac{|(t_2 -\tau_j)- (t_1-\tau_i)|}{L_q}\right].
\end{eqnarray}
We can also obtain the covariance of image $i$ measured at two times $t_1$ and
$t_2$:
\begin{eqnarray}
\textnormal{Cov}\left[y_i(t_1), y_i(t_2)\right]
&=& \mathds{E}\left[\left(y_i(t_1) - m_i\right)\left(y_i(t_2) - m_i\right)\right]\\
&=& \mathds{E}\left[\left(q(t_1 - \tau_i) + \mu_i(t_1)\right)
\left(q(t_2 - \tau_i) + \mu_i(t_2)\right)\right]\\
&=& \mathds{E}\left[q(t_1 - \tau_i)q(t_2 - \tau_i) + q(t_1 - \tau_i)\mu_i(t_2)
+ \mu_i(t_1)q(t_2 - \tau_i) + \mu_i(t_1)\mu_i(t_2)
\right]\\
&=& S_q\exp
\left[-\frac{|(t_2 -\tau_j)- (t_1-\tau_i)|}{L_q}\right] +
S_i\exp\left[
-\left(\frac{|t_2 - t_1|
}{\ell_i}\right)^\alpha
\right]
.
\end{eqnarray}
The covariance between a data point $Y_i$ and another data point $Y_j$ is
given by the covariance of the underlying noise-free light curves at times $t_i$
and $t_j$, plus a noise term:
\begin{eqnarray}
\textnormal{Cov}\left(
Y_i, Y_j
\right) = \textnormal{Cov}\left(y_{k_i}(t_i), y_{k_j}(t_j)\right)
+ \delta_{ij}\left(a_i\sigma_i\right)^2\label{eq:covariance}
\end{eqnarray}
where $\sigma_i$ is the reported error bar given with the data. The coefficient
$a_i$
in front of $\sigma_i$ is the robustness feature that depends on $\kappa_1$,
$\kappa_2$, and the badness variable for data point $i$. Equation~\ref{eq:covariance}
provides the elements of the covariance matrix in the multivariate normal
sampling distribution.
